---
title: "Accelerometer & Gyroscope Data Analysis - Data Cleaning"
author: "Neil Kutty github.com/sampsonsimpson"
date: "June 24, 2016"
output: html_document
---

# Accelerometer & Gyroscope Data Analysis <br>
## Data Cleaning
<br>
<h4> Step 1.) Loading features data </h4>
<hr>
The first line below reads the features.txt file into variable `features`. In the second line, we use `grep` to create a vector called `featuresLite` identifying the feature names that have 'mean' and 'std' in the names.  Finally, we filter on those feature names only and store it in the variable `featureNames`. 

```{r features, eval=FALSE, include=TRUE}
#load features table
features <- read.table('./UCI HAR Dataset/features.txt')

#create vector for features with 'Mean' and 'Std' 
featuresLite <- grep('-[Mm]ean()|-[Ss]td()', features[,2])

#get just the feature names for 'Mean' and 'Std' filtering on logic above
featureNames <- as.character(features[featuresLite,2])
```
<br>
<h4> Step 2.) Clean up feature names and make descriptive </h4>
<hr>

```{r featureNames, eval=FALSE}
#clean up feature names and make them descriptive
featureNames <- gsub('*-meanFreq\\()*', ' MeanFreq ', featureNames)
featureNames <- gsub('*-mean\\()*', ' Mean ', featureNames)
featureNames <- gsub('*-std\\()*', ' StdDev ', featureNames)
```
<br>
<h4> Step 3.) Read in activity names as `activityLabels` and add column names</h4>
<hr>
Here, we name the ID column of the `activityLabels` table `activityID` so that we have a common column name to later merge the `activityLabels$activity` column with the full dataset in Step 6.) below

```{r activityLabels, eval=FALSE}
#read in activity labels as new table
activityLabels <- read.table('./UCI HAR Dataset/activity_labels.txt')
colnames(activityLabels) <- c('activityID','activity')
```
<br>
<h4> Step 4.) Read in Training set data </h4>
<hr>

```{r trainSet, eval=FALSE}
#Read in the Training set data but get only the 'Mean' and 'Std' data
#  according to our featuresLite logical vector created above
subjectTrain <- read.table('./UCI HAR Dataset/train/subject_train.txt')
xtrain <- read.table('./UCI HAR Dataset/train/X_train.txt')[featuresLite]
ytrain <- read.table('./UCI HAR Dataset/train/Y_train.txt')

#Define column names for the training data
colnames(subjectTrain) <- "subjectID"
colnames(xtrain) <- featureNames
colnames(ytrain) <- "activityID"

#combine the subject, y, and x training sets into one
trainSet <- cbind(subjectTrain,ytrain,xtrain)
```
<br>
<h4> Step 5.) Read in Test set data </h4>
<hr>

```{r testSet, eval=FALSE}
#Read in the Test set data but get only the 'Mean' and 'Std' data
#  according to our featuresLite vector created above
subjectTest <- read.table('./UCI HAR Dataset/test/subject_test.txt')
xtest <- read.table('./UCI HAR Dataset/test/X_test.txt')[featuresLite]
ytest <- read.table('./UCI HAR Dataset/test/Y_test.txt')

colnames(subjectTest) <- "subjectID"
colnames(xtest) <- featureNames
colnames(ytest) <- "activityID"

#combine the subject, y, and x test sets into one
testSet <- cbind(subjectTest,ytest,xtest)
```
<br>
<h4> Step 6.) Combine the Test and Training sets into one dataset </h4>
<hr>
In this step, we combine the trainSet and testSet rows into one table using `rbind()`. We use the `merge()` function and join the new `fullSet`to the `activityLabels` table by the `activityID` common column. The `activityLabels$activity` column is then used for the descriptive activity names.  

We then use `tidyr`::`gather()` to gather the columns and change the table from a wide to a long format.
```{r cleanSet, eval=FALSE} 
#combine the training set and test set into one full set
fullSet <- rbind(trainSet,testSet)

#merge the fullSet dataset with the activityLabels table
fullSet <- merge(activityLabels,fullSet,'activityID')

#gather columns into a long format table to create clean dataset
cleanSet <- gather(fullSet,ncol=4:82)
```
<br>
<h4> Step 7.) Create tidy data set </h4>
<hr>
The final step is to create a tidy dataset from the above created `cleanSet`.  We do this using the `tidyr` package and chain operation `%>%`.
<br>
 1.) First we `separate()` the feature column into the three delimited parts
  and labeling the new cols appropriately: 'feature','measure','XYZ'.
<br>
 2.) The next step is to `group_by()` the columns we want to summarise by. (Here, we drop the direction XYZ column so that we only deal with the average of values for all directions for a subject for a particular measurement)
<br>
 3.) The third tidying is to `summarise()` the table to the mean of the values for the features we are grouping by.
<br>
 4.) Finally we `spread()` the dataset by the newly created measure column and cleanup the table before writing it to a file.

```{r tidySet, eval=FALSE}
# separating the feature column into the three delimited parts
#    and labeling the new cols appropriately: 'feature','measure','XYZ'
# grouping by all relevant columns
# summarising by the group above for the average of all values
# spreading the dataset by the newly created 'measure' column
tidySet <- cleanSet %>%
    separate(key, into=c('feature','measure','XYZ'), sep = ' ') %>%
    group_by(feature, activityID, activity, subjectID, XYZ, measure) %>%
    summarise(Value = mean(value)) %>%
    spread(measure, Value)

#clean XYZ column
tidySet$XYZ <- gsub('-','',tidySet$XYZ)
#convert to data frame    
tidySet <- as.data.frame(tidySet)
#save tidySet as Rdata file
save(tidySet, file='tidySet.Rdata')
write.table(tidySet, file='tidySet.txt')
```
